### 2021-02-20 manuscript.Rmd
Add Gram-stain supplemental figure and subject-level models.

---
#title: "Distance from Patient Defines Hotspots for Multidrug-Resistant Organisms in the Healthcare Environment"
#author: "Brendan J. Kelly, MD, MS"
bibliography: bib/references.bib
csl: csl/journal-of-clinical-microbiology.csl
output: html_document
css: style_helvetica_double.css
---

@@

## Title  
Distance from Patient Defines Hotspots for Multidrug-Resistant Organisms in the Healthcare Environment

@@

* __Background__: Multidrug-resistant organisms (MDROs) colonizing the healthcare environment have been shown to contribute to risk for healthcare-associated infections (HAIs), with adverse effects on patient morbidity and mortality. We sought to define how the distance from patient and wastewater sources of bacterial contamination define MDRO hotspots in the healthcare environment.

@@

* __Statistical Methods__: Data were organized using R statistical software version 3.6.1 [@R_Core_Team2018-xo], and plots generated using the “ggplot2” package [@Wickham2016-eh]. We examined the relationship between distance from the patient (towards the wastewater sites) and the probability of MDRO detection using a gaussian process model (exponentiated-quadratic kernel) for distance, with models fit separately for each of 11 detected MDROs. The same gaussian process model evaluated the relationship between distance and the detection of MDRO genera. We used a mixed-effects model with random effects for genus and sampling site to evaluate the relationship between sequence-based detection of bacterial DNA contamination and risk for MDRO detection. Models were fit using Stan Hamiltonian Monte Carlo (HMC) version 2.21, via the "brms" package with default, weakly-informative priors [@Carpenter2017-ec; @Burkner2017-yr]. Prior predictive modeling was performed, and models were fit with 4 chains of 1000 iterations, confirmed with HMC diagnostics (no divergent iterations, Rhat statistic < 1.1 for all parameters, and E-BFMI > 0.2) [@McElreath2016-sf; @Gabry2019-le; @Gelman2020-yu]. We examined parameter distributions at 50%, 80%, and 95% posterior credible intervals to understand the relationship between exposure and outcome variables.

@@

* __Availability of Data__: Sequence data has been uploaded to the National Center for Biotechnology Information's Sequence Read Archive (NCBI SRA): PRJNA689547. Data, analysis scripts, and model code are available at https://github.com/bjklab/Distance-from-Patient-Defines-Hotspots-for-Multidrug-Resistant-Organisms.

@@

* __Impact of Distance from Patient on MDRO Detection by Bacterial Culture__: We related the probability of MDRO detection by selective culture to the distance from the patient and towards the wastewater sites, using mean distances from each of the three in-room sites sampled at each sampling event. **Figure 1** depicts the impact of distance from patient on each of the 11 MDROs encountered in across our cohort. The probability of detecting ESBL and CR Gram-negative organisms, including _Enterobacterales_, _Acinetobacter_, and _Pseudomonas_, increased with distance from the patient. In contrast, _Clostridioides difficile_ and MRSA were more likely to be detected in close proximity to the patient, and less likely to be detected in the wastewater sites. Of note, VRE was distributed more evenly across all areas.

@@

__Figure 1: Impact of distance from the patient on MDRO detection in the healthcare environment.__ Distance from the head of the patient's bed and towards the wastewater sites is shown on the horizontal axis. The vertical axis depicts the probability of MDRO detection by selective culture for each of 11 detected MDROs according to a logistic regression model incorporating a gaussian process for distance. The black line shows the best estimate, and shades of red indicate 50%, 80%, and 95% posterior credible intervals. ESBL = extended-spectrum beta-lactamase; CR = carbapenem-resistant; VRE = vancomycin-resistant _Enterococcus_; MRSA = methicillin-resistant _S. aureus_.

@@

__Figure 2: Impact of distance from the patient on bacterial DNA deposition in the healthcare environment.__ Distance from the head of the patient's bed and towards the wastewater sites is shown on the horizontal axis. The vertical axis depicts the probability of bacterial genus detection by 16S rRNA gene sequencing for genera corresponding to each of 11 detected MDROs according to a logistic regression model incorporating a gaussian process for distance. The black line shows the best estimate, and shades of red indicate 50%, 80%, and 95% posterior credible intervals.

@@

 <!-- 
 <!-- -----  
\newpage
<P style = "page-break-before: always">

### Supplemental Figures: -->

>>>>

---
#title: "Healthcare Micro-Environments Define Multidrug-Resistant Organism Persistence"  
#author: "Brendan J. Kelly, MD, MS"  
bibliography: bib/references.bib
csl: csl/journal-of-clinical-microbiology.csl
output: html_document
css: style_helvetica_double.css
---

@@

## Title  
Healthcare Micro-Environments Define Multidrug-Resistant Organism Persistence

@@

* __Background__: Multidrug-resistant organisms (MDROs) colonizing the healthcare environment have been shown to contribute to risk for healthcare-associated infections (HAIs), with adverse effects on patient morbidity and mortality. We sought to define how the distance from patient and wastewater sources of bacterial contamination define MDRO contamination and persistence in the healthcare environment.

@@

* __Statistical Methods__: Data were organized using R statistical software version 3.6.1 [@R_Core_Team2018-xo], and plots generated using the “ggplot2” package [@Wickham2016-eh]. We examined the relationship between distance from the patient (towards the wastewater sites) and the probability of MDRO detection using a Gaussian process model (exponentiated-quadratic kernel) for distance, with models fit separately for each of 11 detected MDROs. The same Gaussian process model evaluated the relationship between distance and the detection of MDRO genera. We also evaluated mixed-effects model incorporating a random effect for subjects to account for clustering of longitudinal data. To understand the relationship between bacterial DNA contamination and risk for MDRO detection, we used a mixed-effects model with random effects for genus and sampling site. Models were fit using Stan Hamiltonian Monte Carlo (HMC) version 2.21, via the "brms" package with default, weakly-informative priors [@Carpenter2017-ec; @Burkner2017-yr]. Prior predictive modeling was performed, and models were fit with 4 chains of 1000 iterations, confirmed with HMC diagnostics (no divergent iterations, Rhat statistic < 1.1 for all parameters, and E-BFMI > 0.2) [@McElreath2016-sf; @Gabry2019-le; @Gelman2020-yu]. We examined parameter distributions at 50%, 80%, and 95% posterior credible intervals to understand the relationship between exposure and outcome variables.

@@

* __Availability of Data__: Sequence data has been uploaded to the National Center for Biotechnology Information's Sequence Read Archive (NCBI SRA): PRJNA689547. Data, analysis scripts, and model code are available at https://github.com/bjklab.

@@

* __Impact of Distance from Patient on MDRO Detection by Bacterial Culture__: We related the probability of MDRO detection by selective culture to the distance from the patient and towards the wastewater sites, using mean distances from each of the three in-room sites sampled at each sampling event. **Figure 1** depicts the impact of distance from patient on each of the 11 MDROs encountered in across our cohort. The probability of detecting ESBL and CR Gram-negative organisms, including _Enterobacterales_, _Acinetobacter_, and _Pseudomonas_, increased with distance from the patient. In contrast, _Clostridioides difficile_ and MRSA were more likely to be detected in close proximity to the patient, and less likely to be detected in the wastewater sites. Of note, VRE was distributed more evenly across all areas. Given the patterns of MDRO detection observed at the genus level, we also examined the relationship between aggregate Gram-negative or Gram-positive MDROs and the distance from patient (**Supplemental Figure 1**). This analysis confirmed the substantial increase in Gram-negative MDRO detection with increasing distance from the patient (and towards the wastewater site), with the opposite pattern observed among Gram-positive MDROs.

@@

__Figure 1: Impact of distance from the patient on MDRO detection in the healthcare environment.__ Distance from the head of the patient's bed and towards the wastewater sites is shown on the horizontal axis. The vertical axis depicts the probability of MDRO detection by selective culture for each of 11 detected MDROs according to a logistic regression model incorporating a Gaussian process for distance. The black line shows the best estimate, and shades of red indicate 50%, 80%, and 95% posterior credible intervals. ESBL = extended-spectrum beta-lactamase; CR = carbapenem-resistant; VRE = vancomycin-resistant _Enterococcus_; MRSA = methicillin-resistant _S. aureus_.

@@

__Figure 2: Impact of distance from the patient on bacterial DNA deposition in the healthcare environment.__ Distance from the head of the patient's bed and towards the wastewater sites is shown on the horizontal axis. The vertical axis depicts the probability of bacterial genus detection by 16S rRNA gene sequencing for genera corresponding to each of 11 detected MDROs according to a logistic regression model incorporating a Gaussian process for distance. The black line shows the best estimate, and shades of red indicate 50%, 80%, and 95% posterior credible intervals.

@@

### Supplemental Figures:

#### Supplemental Figure 1:

```{r sfig1, eval=TRUE, autodep=TRUE, fig.align='center', out.width='60%'}

knitr::include_graphics(path = "./figs/p_cx_gp_gram_subject.svg")

__Supplemental Figure 1: Impact of distance from the patient on aggregate Gram-negative and Gram-positive MDROs in the healthcare environment.__ Distance from the head of the patient's bed and towards the wastewater sites is shown on the horizontal axis. The vertical axis depicts the probability of MDRO detection by selective culture for Gram-negative or Gram-positive MDROs in aggregate, according to a logistic regression model incorporating a Gaussian process for distance. The black line shows the best estimate, and shades of red indicate 50%, 80%, and 95% posterior credible intervals.
